\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students

### 2023-09-17 phd-cs-us.tex
Updates from Overleaf

Then applications will be reviewed by a
\textbf{PhD admission committee} that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the committee depends on the department size. At GMU, the PhD committee typically has 15--20 faculty, and each committee member is assigned from 25--30 applications. Note GMU and UMass (and probably many other schools) have a separate committee for the MS program.

@@

It's important to aim for the top venues in your field. You can
find top CS conferences at places such as \href{https://csrankings.org}{CSRankings}, which ranks CS programs based on how their faculty publish at top \emph{conferences}. Local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty.

@@

For Vietnamese students, it's worth mentioning the \href{https://vef2.org/}{VEF 2.0 program}, which has helped many good students in gaining acceptance to top-tier PhD programs in the US. VEF 2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), the impact of these reference letters can be very positive as they have helped numerous VEF 2.0 students getting into top CS PhD programs in the US.

@@

International students not familiar with US universities often put them into \emph{two} bins:  (i) very top schools that they dream about such as Stanford, MIT, Princeton, Harvard and (ii) everything else.  In many cases, they use resources such as US News, but their ranking is \emph{highly} questionable~\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes these students rank CS programs using the reputations non-CS programs such as medical, math, or physics.
In some cases they rank universities based on popular states they know in the US, e.g., California and New York.  Clearly, there are so many thing wrong with these methods.

@@

\bibliographystyle{abbrv}
\bibliography{phd-cs-us.bib}
\newpage

\section{Virtual Research Opportunities Beyond Physical Boundaries (written by Didier)}
\didi{In my experience, prospective PhD students (from Rwanda) are often unaware of the information provided in this document throughout their undergraduate. For this reason they often find themselves unprepared for PhD admission process by the time they finish undergraduate and have to spend a year or more to improve their application profiles. I think it would be beneficial to expand the document to also target students who are not yet ready for the application process to help them improve on all criteria considered during admission. I drafted the short essay on how to gain research experience remotely in order to improve their research profile and potentially get LOR from experts (hopefully faculty)}\\
In the realm of computer science, research experience plays a crucial role in securing admissions to top-tier Ph.D. programs. However, students from underrepresented small colleges, both in the United States and internationally, often encounter limited research opportunities within their institutions. Thankfully, there exists an alternative pathway for these aspiring researchers to gain valuable research experience, even without being physically present at a university.

\textbf{Virtual Research Programs}:
Several universities, organizations, and research institutions offer virtual internships and research programs aimed at providing hands-on research experience. These programs often involve working remotely under the guidance of experienced mentors and collaborating with a team of fellow researchers. For instance, \href{https://docs.google.com/forms/d/1btIwt4HwjyKMOUk-EMy3rbkfWzFxv2lNrMm_zkd0pA4/viewform?edit_requested=true}{UIUC+ Summer Undergraduate Research in Software Engineering}  offers an unpaid remote internship for software engineering students all over the world to collaborate with mentors from University of Illinois at Urbana Champaign. 
Virtual internships offer opportunities to contribute to ongoing research projects, conduct experiments, analyze data, and write research reports. Students can search for such programs through university websites, research institutes, or as of recently chatGPT (or Bard).

\textbf{Online Research Communities and Open Source Contributions}:
Online research communities and platforms offer a wealth of opportunities for aspiring researchers. Platforms like GitHub, GitLab, and Bitbucket host repositories spanning diverse research domains. Students can explore repositories related to their areas of interest, collaborate with other contributors, and actively engage in open source research projects. By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field.

\textbf{Online Conferences and Workshops}:
Attending online conferences and workshops is another way to gain exposure to cutting-edge research and establish connections with experts in the field. Many conferences now provide virtual participation options, enabling students to access research talks, poster sessions, and panel discussions and sometimes access designated chat rooms or networking events where participants can engage with researchers, ask questions, and seek potential research collaborations. It is beneficial to also create profiles academic collaboration platforms such as ResearchGate. By creating profiles, joining relevant research groups, and participating in discussions, students can connect with established researchers, contribute to ongoing projects, and potentially collaborate on publications or research proposals as they provide access

In conclusion, for students lacking research opportunities at their small colleges, the virtual realm offers a wealth of possibilities to gain valuable research experience. Engaging with open source projects, joining online research communities, utilizing academic collaboration platforms, participating in virtual internships, and attending online conferences are all avenues to contribute to research, connect with experts, and enhance one's research profile.

>>>>

Then applications will be reviewed by a
\textbf{PhD admission committee} that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the committee depends on the department size. At GMU, the PhD committee typically has 15--20 faculty, and each committee member is assigned from 25--30 applications. Note GMUU, UMass, and probably many other schools, have a separate committee for the MS program.

@@

It's important to aim for the top venues in your field. You can
find top CS conferences at places such as \href{https://csrankings.org}{CSRankings}, which ranks CS programs based on how their faculty publish at top \emph{conferences} (see more in \S\ref{sec:ranking}). Local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty.

@@

For Vietnamese students, it's worth mentioning the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining acceptance to top PhD programs in the U. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), the impact of these reference letters can be positive as they have helped many students getting into top PhD programs in the US.

@@

International students not familiar with US universities often put them into \emph{two} bins:  (i) very top schools that they dream about such as Stanford, MIT, Princeton, Harvard and (ii) everything else.  In many cases, they use resources such as US News, but their ranking is \emph{highly} questionable~\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes these students evaluate CS programs using the reputations non-CS programs such as medical, math, or physics.
In some cases they rank universities based on popular states they know in the US, e.g., California and New York.  Clearly, there are so many thing wrong with these methods.

@@

\bibliographystyle{abbrv}
\bibliography{phd-cs-us.bib}

% \newpage
% \section{Virtual Research Opportunities Beyond Physical Boundaries (written by Didier)}
% \didi{In my experience, prospective PhD students (from Rwanda) are often unaware of the information provided in this document throughout their undergraduate. For this reason they often find themselves unprepared for PhD admission process by the time they finish undergraduate and have to spend a year or more to improve their application profiles. I think it would be beneficial to expand the document to also target students who are not yet ready for the application process to help them improve on all criteria considered during admission. I drafted the short essay on how to gain research experience remotely in order to improve their research profile and potentially get LOR from experts (hopefully faculty)}\\
% In the realm of computer science, research experience plays a crucial role in securing admissions to top-tier Ph.D. programs. However, students from underrepresented small colleges, both in the United States and internationally, often encounter limited research opportunities within their institutions. Thankfully, there exists an alternative pathway for these aspiring researchers to gain valuable research experience, even without being physically present at a university.

% \textbf{Virtual Research Programs}:
% Several universities, organizations, and research institutions offer virtual internships and research programs aimed at providing hands-on research experience. These programs often involve working remotely under the guidance of experienced mentors and collaborating with a team of fellow researchers. For instance, \href{https://docs.google.com/forms/d/1btIwt4HwjyKMOUk-EMy3rbkfWzFxv2lNrMm_zkd0pA4/viewform?edit_requested=true}{UIUC+ Summer Undergraduate Research in Software Engineering}  offers an unpaid remote internship for software engineering students all over the world to collaborate with mentors from University of Illinois at Urbana Champaign. 
% Virtual internships offer opportunities to contribute to ongoing research projects, conduct experiments, analyze data, and write research reports. Students can search for such programs through university websites, research institutes, or as of recently chatGPT (or Bard).

% \textbf{Online Research Communities and Open Source Contributions}:
% Online research communities and platforms offer a wealth of opportunities for aspiring researchers. Platforms like GitHub, GitLab, and Bitbucket host repositories spanning diverse research domains. Students can explore repositories related to their areas of interest, collaborate with other contributors, and actively engage in open source research projects. By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field.

% \textbf{Online Conferences and Workshops}:
% Attending online conferences and workshops is another way to gain exposure to cutting-edge research and establish connections with experts in the field. Many conferences now provide virtual participation options, enabling students to access research talks, poster sessions, and panel discussions and sometimes access designated chat rooms or networking events where participants can engage with researchers, ask questions, and seek potential research collaborations. It is beneficial to also create profiles academic collaboration platforms such as ResearchGate. By creating profiles, joining relevant research groups, and participating in discussions, students can connect with established researchers, contribute to ongoing projects, and potentially collaborate on publications or research proposals as they provide access

% In conclusion, for students lacking research opportunities at their small colleges, the virtual realm offers a wealth of possibilities to gain valuable research experience. Engaging with open source projects, joining online research communities, utilizing academic collaboration platforms, participating in virtual internships, and attending online conferences are all avenues to contribute to research, connect with experts, and enhance one's research profile.
